Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-03-23 Ben_Manuscripts/hdphmm/Outline.tex
comments on 2.2

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, H, a Dirichlet process 
    (DP) generates distributions over a countably infinite number of 
    probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the concentration 
    parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    \item The concentration parameter expresses one's confidence in H relative to the posterior 
    and is closely related to the number of data observations.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden states
  while simultaneously estimating VAR(1) parameters for each state and the overall 
  state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    as shown in SI.
    \item We refer the interested reader to much more extensive descriptions of 
    this process and its implementation. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  There are many ways one can apply the IHMM algorithm to timeseries data. 
  \begin{itemize}
    \item We begin by trying to identify as many distinct states as possible
    in each solute trajectory and then take advantage of the system's cylindrical
    symmetry in order to cluster the parameter sets into an interpretable number
    of states.
    \item Having a large number of states gives more reliable clustering results.
%    \item We believe the subsequent procedure is best suited for our data
%    because it can clearly distinguish different dynamical behavior, it takes
%    advantage of the system's cylindrical symmetry and represents distinct
%    dynamics in an interpretable number of states.
  \end{itemize}

  We first applied the IHMM algorithm to 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with the center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$ (see Figure~\ref{fig:cartesian_cylinder}).
    \item By using 2 dimensions, rather than a single radial dimension, we
    increase the likelihood of finding more states since a single radial
    mean can be represented by an infinite number of $x$ and $y$ coordinate combinations.
  \end{itemize}

  \begin{figure}
  \centering
  \includegraphics[width=0.2\textwidth]{cartesian_cylinder.pdf}
  \caption{Using the nearest pore center as the origin, we represented the solute's
  location along the pore axis in terms of the $z$ coordinates and their radial distance
  from the pore centers in 2 dimensions, $x$ and $y$.}\label{fig:cartesian_cylinder}
  \end{figure}

@@

  The states predicted by the IHMM are heavily influenced by the Gaussian prior
  placed on $c$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of the $A$ and $\mathbf{e}_t$ matrices do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the means of 
    each state lie within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to be halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the maximum and minimum
    to be 2 standard deviations from the mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

@@

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences and parameters for each state.
  \begin{itemize}  
    \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We averaged the parameters from each iteration recorded after the
    equilibration time point.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    \item Since all of the parameters are multidimensional, we found the 
    equilibration time point of each component and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

  % BJC2: It's probably possible to convert to from x, y, z to r, z but
  % I need to figure out the transformation of A. This approach works though.
  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because the mean in $r$ is 
  a useful clustering variable.
  \begin{itemize}
    \item It will also be clearer to analyze the final parameter set in
    terms of $r$ rather than $x$ and $y$.
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
    \item In order to keep the same states found when we applied the algorithm
    in 3D, we reapplied the IHMM to the cylindrical trajectories with a fixed 
    state sequence, only allowing inference on the VAR(1) parameters. 
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
   \item For each solute studied, we identified between 200-300 independent sets of
   parameters. % BJC2: just a guess for now
   \item Many of these states exhibit very similar dynamical behavior except their
   mean levels are different, especially in the axial direction where solutes could
   get trapped along a broad and continuous range of $z$ coordinates.
   \item Since we do not know the number of states with shared dynamical behavior
   beforehand, we used a non-parametric Bayesian Gaussian mixture model in order
   to group them. % cite scikit-learn  
%   \item This is necessary for us to perform any kind of mechanistic speculation.
  \end{itemize}

@@

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states.
  \end{itemize}

@@

  We used the IHMM algorithm in order to infer $A$ and $\mathbf{e}_t$ of the 
  clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $\mathbf{e}_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segement of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets.
  \begin{itemize}
    \item We drew state sequences with transition probabilities weighted by the 
    the rows of $T$.
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item After each state transition, we reset the unconditional mean of each 
    state based on the particle's position immediately before the state transition
    occurred.
  \end{itemize}

>>>>

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, H, a Dirichlet process 
    (DP) generates distributions over a countably infinite number of 
    probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the concentration 
    parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    %MRS3: This next line is a bit vauge (expresses confidence, closely related)
   \item The concentration parameter $\gamma$ expresses one's confidence in H relative to the posterior 
    and is closely related to the number of data observations.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden states
  while simultaneously estimating VAR(1) parameters for each state and the overall 
  state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    %MRS3: be a bit more specific about what the criteria were for convergence.
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    as shown in SI.
    \item We refer the interested reader to much more extensive descriptions of 
    %MRS3: be more specific about which parts from the above are in the references
    this process and its implementation. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  %MRS3: be more specific - i.e. describe what are the ways you are going to apply it. 
  There are many ways one can apply the IHMM algorithm to timeseries data. 
  \begin{itemize}
    %MRS3: it's not as many as possible, but identify one's supported by the data, right?
    \item We begin by trying to identify as many distinct states as possible
    in each solute trajectory and then take advantage of the system's cylindrical
    symmetry in order to cluster the parameter sets into an interpretable number
    of states.
    %MRS3: not clear why this is the case?
    \item Having a large number of states gives more reliable clustering results.
%    \item We believe the subsequent procedure is best suited for our data
%    because it can clearly distinguish different dynamical behavior, it takes
%    advantage of the system's cylindrical symmetry and represents distinct
%    dynamics in an interpretable number of states.
  \end{itemize}

  We first applied the IHMM algorithm to 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with the center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$ (see Figure~\ref{fig:cartesian_cylinder}).
    %MRS3: I'm not following the below? I don't know that we WANT to find more states, in my mind doing it this way 
    %the math is simpler this way since we don't have to rewrite things in term of radial diffusion
    %MRS3: we can brainstorm this as needed!
    \item By using 2 dimensions, rather than a single radial dimension, we
    increase the likelihood of finding more states since a single radial
    mean can be represented by an infinite number of $x$ and $y$ coordinate combinations.
  \end{itemize}

  \begin{figure}
  \centering
  %MRS3: does this preserve the z-axis symmmetry? Is it just ensuring
  %that you can't jump that far, or is it limitinig the values that
  %the mean can take? Apologies if it's not clear to me yet.  I'm
  %pretty sure it's just limiting the jump (since y_t = c + f(y_t-1)),
  %but would be good to make that explicit - ``i.e. we need to make sure it doesn't jump too far'')
  \includegraphics[width=0.2\textwidth]{cartesian_cylinder.pdf}
  \caption{Using the nearest pore center as the origin, we represented the solute's
  location along the pore axis in terms of the $z$ coordinates and their radial distance
  from the pore centers in 2 dimensions, $x$ and $y$.}\label{fig:cartesian_cylinder}
  \end{figure}

@@

  The states predicted by the IHMM are heavily influenced by the Gaussian prior
  placed on $c$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of the $A$ and $\mathbf{e}_t$ matrices do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the means of 
    each state lie within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to be halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the maximum and minimum
    to be 2 standard deviations 
    %MRS3: from the mean of what?
    from the mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

@@

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences and parameters for each state.
  \begin{itemize}  
    %MRS3: can you say more which categories converged quickly?
   \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We averaged the parameters from each iteration recorded after the
    equilibration time point.
    %MRS3: averaged and used this as estimate for running models later? 
    %MRS3: other possbility is maximum a posteriori estimate, but for peaked symmetric distributions, it's essentially the same.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    %MRS3: not sure what you mean that they all were multidimensional?
    \item Since all of the parameters are multidimensional, we found the 
    equilibration time point of each component and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

  % BJC2: It's probably possible to convert to from x, y, z to r, z but
  % I need to figure out the transformation of A. This approach works though.
  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because the mean in $r$ is 
  a useful clustering variable.
  %MRS3: because of the radial symmetry, x and y aren't really meaningful alone. 
  \begin{itemize}
    \item It will also be clearer to analyze the final parameter set in
    terms of $r$ rather than $x$ and $y$.
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
    \item In order to keep the same states found when we applied the algorithm
    in 3D, 
    %MRS3: would it be clearer to say you appplied only the inference part of the process on the clustered data?
    we reapplied the IHMM to the cylindrical trajectories with a fixed 
    state sequence, only allowing inference on the VAR(1) parameters. 
    %MRS3:did you cluster the state transitions as well? 
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  %MRS3: Maybe ``independent sets'' -> ``independent states, each with separate VAR(1) parameters?''
  \item For each solute studied, we identified between 200-300 independent sets of
   parameters. % BJC2: just a guess for now
   \item Many of these states exhibit very similar dynamical behavior except their
   mean levels are different, especially in the axial direction where solutes could
   get trapped along a broad and continuous range of $z$ coordinates.
   \item Since we do not know the number of states with shared dynamical behavior
   beforehand, we used a non-parametric Bayesian Gaussian mixture model in order
   %MRS3: group -> cluster
   to group them. % cite scikit-learn  
%   \item This is necessary for us to perform any kind of mechanistic speculation.
  \end{itemize}

@@

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states.
   %MRS3: seems like you should just be able to lump together the IHMM states to generate the IHMM estimate of T?
  \end{itemize}

@@

  %MRS3: ``used the IHMM algorithm'' -> ``used the parameter inference part of the IHMM algorithm''?
  We used the IHMM algorithm in order to infer $A$ and $\mathbf{e}_t$ of the 
  clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $\mathbf{e}_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segement of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets.
  \begin{itemize}
    %MRS3: not sure what is meant by ``Weighted by the rows of T''? Should it just be given by T? 
    \item We drew state sequences with transition probabilities weighted by the 
    the rows of $T$.
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item After each state transition, we reset the unconditional mean of each 
    state based on the particle's position immediately before the state transition
    occurred.
  \end{itemize}
